---
title: "Practical Class 6: Biodiversity changes"
author: "François Leroy"
format:
  html:
    toc: true
    toc_float: true
editor: visual
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = F,
  warning = F
)
```

**Content of the workshop of today**

We will continue to work with the North American BBS that we worked with already. Last week, we assessed several metrics of biodiversity for the year 2019. However, we can wonder: how has bird biodiversity changed through time (*e.g.* from 1980 to 2021)?

To answer this question, we need **time-series** *i.e.*, data which have been harvested for several years at the same location. This is the case of the BBS: each road has been visited once a year for a very long time. Today, we will work with data from 1990 to 2021 and assess how had biodiversity changed through this period for the different BBS roads.

Also, when working with time-series, one thing to consider is how the repeated samples have been made through the years. Were they sampled for the same amount of time? By the same number of volunteers...? This is what is often called in ecology the *sampling effort* or the *study design*. Fortunately, the BBS data collection is standardized, which means that the volunteers have to follow a protocol when counting the birds. Thus, we can compare the data through time without having to account for the sampling effort.

# Assessing temporal change of diversity

There are two ways to assess changes of biodiversity:

1.  We can simply measure biodiversity (for instance with species richness) and then see how this metric has changed through time. This is called a **temporal trend**.

2.  We can use metrics which are representing temporal dynamic of biodiversity. Here, we will study the three metrics responsible of the change of species richness: *colonization*, *extinction* and *persistence*.

# Temporal changes of species richness

We will first see how species richness has changed through time for the different BBS roads.

```{r}
rm(list = ls())
## A bit of data management. This is the same as we did last week. 
library(readr)
library(tidyr)
library(dplyr)
## Get the file names
file_list <- list.files("data/50-StopData/1997ToPresent_SurveyWide/",
                        full.names = T)
## Create the dataframe that will contain the data
species <- data.frame()
## Loop over each file
for(i in 1:length(file_list)){
  ## Load the file and temporary store it
  d <- read_csv(file_list[i])
  ## Bind to the species dataframe
  species <- rbind(species, d)
  ## Remove the temporary data
  rm(d)
}

## Also, we will keep only the roads which have data in both 2000 and 2021
species <- 
  species %>% 
  group_by(CountryNum, StateNum, Route) %>% 
  filter(any(Year == 2000)) %>%
  filter(any(Year == 2021)) %>% 
  filter(Year %in% c(2000:2021)) %>% 
  ungroup()
```

```{r}
## Unlike last week, today we will work on the entire dataset, and not only year 2019.
## Let's compute species richness for each road and each year using the RouteDataID column.
species_richness <- 
  species %>% 
  group_by(RouteDataID) %>%
  summarize(sr = length(unique(AOU)),        ## Compute species richness
            CountryNum = unique(CountryNum), ## Keep the route info
            StateNum = unique(StateNum),     ## Keep the route info
            Route = unique(Route),           ## Keep the route info
            Year = unique(Year))             ## Keep the year
head(species_richness)
```

Let's represent the values of species richness through time for a single route. It is important to note that the unique identifier of a road is composed by the combination of the three columns: `CountryNum + StateNum + Route`.

```{r}
library(ggplot2)
## Let's pick up randomly one road ####
species_richness %>% 
  filter(CountryNum == 840,
         StateNum == "35",
         Route == "005") %>% 
  #### and plot the value of species richness against time
  ggplot(aes(x = Year, y = sr))+
  geom_point()+
  theme_bw()
```

Well, it seems like there is a linear relationship between species richness and time isn't it? If you remember the class of last week, you know that we can fit a simple linear regression to explain how species richness is *"driven"* by the time (in other words how it behaves through time):

$$SR = \beta_1t + \beta_0$$

**Try to explain how** $\beta_1$ **and** $\beta_0$ **link the species richness with time**

Hereafter, we will fit simple linear regressions of species richness as a function of time for each road of the BBS. 

```{r}
## Now let's fit a linear regression between species richness and time for each road. We could use a for loop but R as a cool function: do()
sr_trend <-
  species_richness %>% 
  group_by(CountryNum, StateNum, Route) %>%  ## those 3 columns represent the route id
  do({
      mod <- lm(sr ~ Year, data = .)                               #### fit the linear model 
      data.frame(slope = coef(mod)[2],                             #### extract the slope
                 intercept = coef(mod)[1],                         #### extract the intercept
                 pval = (summary(mod)$coefficients[2,"Pr(>|t|)"])) #### extract the p-value
      
    })

# We can display the distribution of the slope values using
hist(sr_trend$slope)
```

**What can we say about the trends of species richness from 2000 to 2021 for the BBS roads?**

**Using the script of last week, display on a map the BBS roads with the color indicating the temporal trend of species richness (exemple below)**
*Hint:* in order to set the color gradient from red to blue passing by white, you should use the function `scale_color_gradient2()`. Use `?scale_color_gradient2()` to know more.

```{r, echo=F}
library(sf)
routes <- read.csv(file = "data/routes.csv")
## Create your own CRS for North America
crs_NA <- "+proj=laea +lon_0=-68.91 +lat_0=1.28 +datum=WGS84 +units=m +no_defs"
## Transform the dataframe into an sf object
routes_shp <- st_as_sf(routes,
                       coords = c("Longitude", "Latitude"), ## specify the columns containing the longitude and latitude
                       crs = 4326) ## specify the CRS of the data
## Set the crs to the one created above
routes_shp <- routes_shp %>% st_transform(crs = crs_NA)
# install.packages("rnaturalearthdata")
library(rnaturalearth)
countries_shp <-
  ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(continent == "North America") %>%
  st_transform(crs = crs_NA)
## The class() of the coulmns StateNum and Route must be the same between the 2 datasets:
sr_trend <- mutate(sr_trend, StateNum = as.numeric(StateNum), Route = as.numeric(Route))
## Merge the 2 datasets using left join
routes_shp <- left_join(routes_shp, sr_trend,
                        by = c("CountryNum", "StateNum", "Route"))
## Map
ggplot()+
  geom_sf(data = countries_shp)+
  geom_sf(data = na.omit(routes_shp), aes(color = slope))+
  scale_color_gradient2(low = "red", midpoint = 0, mid = "white", high = "blue")+
  ggtitle("Change of species richness between 2000-2021 for each BBS road")+
  theme_bw()+
  theme(plot.title = element_text(hjust = .5))
```

# Assessing colonization, extinction and persistence

```{r}

```

