---
title: "Drivers of species distribution"
author:
  - name: 'Florencia Grattarola'
    orcid: 0000-0001-8282-5732
  - name: 'Friederike Wölke'
    orcid: 0000-0001-9034-4883
  - name: 'Gabriel Ortega'
    orcid: 0000-0002-0516-5694
format: 
  html:
    toc: true
    toc-location: right
    smooth-scroll: true
    html-math-method: katex
    code-fold: true
self-contained: true
editor: source
editor_options: 
  chunk_output_type: console
---

## Peru: drivers of amphibians richness 🐸

1.    Get a map of Peru
2.    Get 5 different drivers (`raster`) 
3.    Process all the layers to the same resolution and the extent of Peru
4.    Visualise predictors 
5.    Calculate the correlation between the different predictors (and compare with biodiversity) 


### Data download

Please download the following file and store it on you `data` folder:  

-   `amphibians_grid.rds` 

### Libraries 

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

library(pacman)
pacman::p_load(sf,  # Manage spatial data as simple features dataframes
               rnaturalearth, # useful base maps
               tidyverse) # install the package if you haven't already

sf_use_s2(FALSE) # switch spherical geometry off

pacman::p_load(terra, # Work with rasters and vectorial spatial data
               tmap, # Cool maps
               GGally, # Cool correlation maps
               geodata, # Get data from multiple sources
               MODIStsp) # Get data from MODIS
```

### 1.    Get a map of Peru

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

peru <- ne_countries(
  country = "Peru",
  returnclass = "sf" # We want an object of class sf
) 
```

Let’s see how it looks.

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

ggplot() + 
  geom_sf(data=peru, fill='white') +
  coord_sf()
```


To get a planar projection of the area (visit <https://projectionwizard.org/>).


```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

proj_peru <- 'PROJCS["ProjWiz_Custom_Transverse_Cylindrical_Equal_Area",
 GEOGCS["GCS_WGS_1984",
  DATUM["D_WGS_1984",
   SPHEROID["WGS_1984",6378137.0,298.257223563]],
  PRIMEM["Greenwich",0.0],
  UNIT["Degree",0.0174532925199433]],
 PROJECTION["Transverse_Cylindrical_Equal_Area"],
 PARAMETER["False_Easting",0.0],
 PARAMETER["False_Northing",0.0],
 PARAMETER["Central_Meridian",-74.7949219],
 PARAMETER["Scale_Factor",1.0],
 PARAMETER["Latitude_Of_Origin",0.0],
 UNIT["Meter",1.0]]'
```


### 2.    Get 5 different drivers (`raster`) 

-   Annual mean temperature (**WorldClim BIO1**)
-   Precipitation seasonality (**WorldClim BIO15**)
-   Elevation (**SRTM**)
-   Land-cover (**MODIS Terra MCD12Q1**)
-   Human footprint index (**SEDAC**)

When working with rasters, there are three basic procedures we can recommend you:

1.  **Crop**: Cut the rasters to the geographic `extent` you need. This is a square cut.
2.  **Reproject**: Make sure that all your rasters are on the same `projection`. But be aware that by reprojecting, you are introducing an additional source of error.
3.  **Mask**: Set to `NA` all the cells outside your polygon(s) of interest.


We will create a variable with the spatial extent of Peru to use it to crop the raster data.

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

round(st_bbox(peru),0) # Check the extent of the country (round values)
round(st_bbox(peru),0) + c(-1,-1,1,1) #This way we add space around the country
```

Now let's create an object with the extent.

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

extent <- round(st_bbox(peru),0) + c(-1,-1,1,1) # Bounding box plus 1 degree
```


We will also set the path were we are going to save data. 

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

geodata_path("data/")
```

This is required to use the package `geodata`. From now on, every time you try to get a dataset with `geodata` functions, the package will check if the data already exists in your data folder and load it from there.


#### Drivers: climate

Let's first download the **WorldClim** data (version 2.1 climate data for 1970-2000):

-   Annual mean temperature (`bio1`)  
-   Annual precipitation (`bio15`)  

You can download these data from <https://www.worldclim.org/data/worldclim21.html> <br> or use the package `geodata`.
We will use the function `worldclim_country()` from `geodata`.

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

tavg <- worldclim_country(var = "tavg",country = "PER") # PER is the three letters ISO code for PERU
tavg <- crop(tavg, extent)
```

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

tavg
```

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

tavg
```

Now get the `bio` data from Worldclim, and filter `bio12`.

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

bio <- worldclim_country(var = "bio", country = "PER")
prec <- bio$wc2.1_30s_bio_12

prec <- crop(prec, extent)
prec
```

#### Drivers: elevation 

We will use the function `elevation_30s()` from `geodata`. These data come from the Shuttle Radar Topographic Mission and are aggregated to 1km resolution.

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

elev <- elevation_30s(country = "PER")

elev <- crop(elev, extent)

elev
```

#### Drivers: land-cover 

Now, let's get the landcover data. For this we will use the package `MODIStsp`.

::: {.callout-important}
## `MODIStsp` has not been maintained
The package may not be working properly.
:::

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

MODIStsp_get_prodnames()
```


To download **MODIS** data through the NASA *http* server, we need to create a profile at <https://urs.earthdata.nasa.gov/home> to get a user and password.

Today, I'll present you an example and provide you with the data already processed, but you can do this at home.

We will use the [Land Cover Type Yearly L3 Global 500m](https://lpdaac.usgs.gov/products/mcd12q1v006/), and download data for **Peru** from the year **2020**.


With `MODIStsp_get_prodlayers()` you can see all the layers of the product:

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

MODIStsp_get_prodlayers("MCD12Q1", prodver = "061")$bandnames
```

There are five different land cover classification schemes, we will be using the primary land cover scheme (**LC1**) which identifies 17 classes defined by the [IGBP](http://www.igbp.net/) (International Geosphere-Biosphere Programme), including 11 natural vegetation classes, 3 human-altered classes, and 3 non-vegetated classes.

```{r}
#| echo: true
#| eval: false
#| code-block-border-left: true

MODIStsp(
  gui = FALSE,
  out_folder = "data/",
  out_folder_mod = "data/",
  selprod = "LandCover_Type_Yearly_500m (MCD12Q1)",
  prod_version = "061",
  bandsel = "LC1",
  sensor = "Terra",
  user = your_user, # your username for NASA http server
  password = your_pass, # your password for NASA http server
  start_date = "2020.01.01",
  end_date = "2020.12.31",
  verbose = TRUE,
  bbox = extent, # bbox covering Peru
  spatmeth = "bbox",
  out_format = "GTiff",
  compress = "LZW",
  out_projsel = "User Defined",
  output_proj = 4326,
  delete_hdf = TRUE,
  parallel = TRUE
)
```

Let's read the layer and process it.

```{r}
#| echo: true
#| eval: true
#| code-block-border-left: true

land <- rast("data/MCD12Q1.061_LC_Type1_doy2020001_aid0001.tif")

land <- crop(land , extent)

land
```

#### Drivers: anthropogenic

Again using the package `geodata`.

```{r}
#| echo: true
#| eval: true

footp <- footprint(year = 2009)

footp <- crop(footp, extent)

footp
```

### 3.    Process all the layers to the same resolution and the extent of Peru

Rasters should have the same spatial extent and resolution...
If they don't match... what should we do?

```{r}
#| echo: true
#| eval: true

compareGeom(tavg, prec, elev, footp, land, stopOnError = F, messages = T, res = T)
```

```{r}
#| echo: true
#| eval: true

st_bbox(tavg)
st_bbox(prec) 
st_bbox(elev) 
st_bbox(footp)
st_bbox(land)

#Compare with the country extent
st_bbox(peru)
```

We will use the package `terra` to process the rasters.\
First we need to create a new raster to use as a **template**.

::: {.callout-tip collapse="true"}
## Bug in `terra`

We could do:

```
r <- rast(res = 10000,
          ext = extention, # or even `ext(vect(peru))`
          crs = "epsg:4326") 
```

But there's a bug :O\
I reported it <https://github.com/rspatial/terra/issues/1618>
:::

```{r}
#| echo: true
#| eval: true

r <- rast(res = 10000, # Target cell size
          ext(vect(st_transform(peru, proj_peru))), # Extent of the raster
          crs = proj_peru) %>% # Useful planar projection
  project(., "epsg:4326") # Projection of our rasters

r
```

We will `resample()` all of the rasters to our target resolution.

```{r}
#| echo: true
#| eval: true

tavg <- resample(tavg, r, method = "bilinear") # for continuous data!
prec <- resample(prec, r, method = "bilinear")
elev <- resample(elev, r, method = "bilinear")
footp <- resample(footp, r, method = "near") # for categorical data!
land <- resample(land, r, method = "near") 
```

And we can compare the rasters after resampling.

```{r}
#| echo: true
#| eval: true

compareGeom(tavg, prec, elev, footp, land, stopOnError = F, messages = T, res = T)
```

And finally we will use `mask()` to intersect the layer with Peru.

```{r}
#| echo: true
#| eval: true

tavg <- mask(tavg, vect(peru))
prec <- mask(prec, vect(peru))
elev <- mask(elev, vect(peru))
footp <- mask(footp, vect(peru))
land <- mask(land, vect(peru))
```

### 4.    Visualise predictors

We will use the package `tmap`.

```{r}
#| echo: true
#| eval: true

tmap_mode(mode = "view")
```

#### Annual mean temperature (WorldClim BIO1)

```{r}
#| echo: false
#| eval: true

tm_shape(mean(tavg)) +
  tm_raster(palette = "OrRd",
            midpoint = NA, 
            style = "cont", n = 10)
```


#### Precipitation seasonality (WorldClim BIO15)

```{r}
#| echo: false
#| eval: true

tm_shape(mean(prec)) +
  tm_raster(palette = "RdBu", 
            midpoint = NA, 
            style = "cont", n = 10) +
  tm_view(bbox = extent)
```

---

#### Elevation (WorldClim)

```{r}
#| echo: false
#| eval: true

tm_shape(elev) +
  tm_raster(palette = "BrBG", 
            midpoint = NA, 
            style = "cont", n = 10) +
  tm_view(bbox = extent) +
  tm_layout(legend.position = 'left')
```

#### Land-cover (MODIS Terra MCD12Q1)

We need to make sure values are factors.

```{r}
#| echo: true
#| eval: true

names(land)
```

```{r}
#| echo: true
#| eval: true

land <- as.factor(land$MCD12Q1.061_LC_Type1_doy2020001_aid0001)
```

Then we can add values to the levels. 

```{r}
#| echo: true
#| eval: true

levels(land[[1]]) <- data.frame(
  ID = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17),
  label = c(
    "Evergreen Needleleaf Forest",
    "Evergreen Broadleaf Forests",
    "Deciduous Needleleaf Forests",
    "Deciduous Broadleaf Forests",
    "Mixed Forests",
    "Closed Shrublands",
    "Open Shrublands",
    "Woody Savannas",
    "Savannas",
    "Grassslands",
    "Permanent Wetlands",
    "Croplands",
    "Urban and Built-up Lands",
    "Cropland/Natural Vegetation Mosaics",
    "Permanent Snow and Ice",
    "Barren",
    "Water Bodies"
  )
)
```

Please bear in mind that your landcover layer might not have the same classes/levels as mine. Check here the classification values: [MCD12_User_Guide_V6](https://lpdaac.usgs.gov/documents/101/MCD12_User_Guide_V6.pdf) (see page 7 Classification Legends).


```{r}
#| echo: false
#| eval: true

tm_shape(land) +
  tm_raster(palette = "Set1", 
            style = "cat") +
  tm_view(bbox = extent) +
  tm_layout(legend.position = 'left')
```

#### Human footprint index (SEDAC)

```{r}
#| echo: false
#| eval: true
#| layout-ncol: 1

tm_shape(footp) +
  tm_raster(palette = "viridis", style = "cont") +
  tm_view(bbox = extent)
```

### 5.    Calculate the correlation between the different predictors (and compare with biodiversity) 

Let's get values per grid-cell.

```{r}
#| echo: true
#| eval: true

names(tavg)
```

Let's see all the values together.

```{r}
#| echo: true
#| eval: true

predictors <- tibble(
  "tavg" = values(tavg$PER_wc2.1_30s_tavg_1) %>% as.vector(),
  "prec" = values(prec$PER_wc2.1_30s_prec_1) %>% as.vector(),
  "elev" = values(elev) %>% as.vector(),
  "land" = values(land) %>% as.vector(),
  "footp" = values(footp) %>% as.vector()
) %>% 
  mutate(
    across(everything(), ~replace_na(.x, NA))
  )
```

```{r}
#| echo: true
#| eval: true

summary(predictors)
```

Let's see how these values correlate.

```{r}
#| echo: true
#| eval: true
#| warning: false

GGally::ggpairs(predictors, cex = 0.1)
```

What can this plot tell you?

### EXTRA: Can we see any pattern with the hotspots of amphibians species richness?

Let's read the data, plot it and see the patterns

```{r}
#| echo: true
#| eval: true

amphibians_grid <- readRDS("data/amphibians_grid.rds") %>%
  st_transform(., 4326)
```


```{r}
#| echo: false
#| eval: true
#| layout-ncol: 1

tm_shape(amphibians_grid) +
  tm_fill("SR", palette = "YlOrBr") +
  tm_view(bbox = extent)
```

Species richness vs mean annual temperature

```{r}
#| echo: false
#| eval: true
#| layout-ncol: 2

tm_shape(amphibians_grid) +
  tm_fill("SR", palette = "YlOrBr") +
  tm_view(bbox = extent)

tm_shape(mean(tavg)) +
  tm_raster(palette = "OrRd", midpoint = NA, style = "cont", n = 10) +
  tm_view(bbox = extent)
```

Species richness vs precipitation seasonality

```{r}
#| echo: false
#| eval: true
#| layout-ncol: 2

tm_shape(amphibians_grid) +
  tm_fill("SR", palette = "YlOrBr") +
  tm_view(bbox = extent)

tm_shape(mean(prec)) +
  tm_raster(palette = "RdBu", midpoint = NA, style = "cont", n = 10) +
  tm_view(bbox = extent)

```


Species richness vs elevation

```{r}
#| echo: false
#| eval: true
#| layout-ncol: 2

tm_shape(amphibians_grid) +
  tm_fill("SR", palette = "YlOrBr") +
  tm_view(bbox = extent)

tm_shape(elev) +
  tm_raster(palette = "BrBG", midpoint = NA, style = "cont", n = 10) +
  tm_view(bbox = extent)
```


Species richness vs land-use

```{r}
#| echo: false
#| eval: true
#| layout-ncol: 2

tm_shape(amphibians_grid) +
  tm_fill("SR", palette = "YlOrBr") +
  tm_view(bbox = extent)

tm_shape(land) +
  tm_raster(palette = "Set1", style = "cat") +
  tm_view(bbox = extent)
```

Species richness vs human footprint

```{r}
#| echo: false
#| eval: true
#| layout-ncol: 2

tm_shape(amphibians_grid) +
  tm_fill("SR", palette = "YlOrBr") +
  tm_view(bbox = extent)

tm_shape(footp) +
  tm_raster(palette = "viridis", style = "cont") +
  tm_view(bbox = extent)
```

We are done! No it's **your turn** :)

## France: drivers of trees richness

1.    Get a map of France
2.    Get 5 different drivers (`raster`) 
3.    Process all the layers to the same resolution and the extent of France
4.    Visualise predictors 
5.    Calculate the correlation between the different predictors (and compare with biodiversity) 



### 1.    Get a map of France

This might be useful: <http://bboxfinder.com/>

```{r}
#| echo: true
#| eval: true

france <- ne_countries(country = "France", scale=50, returnclass = "sf") 
bbox_france <- st_bbox(c(xmin = -6.855469, xmax = 12.128906, 
          ymax = 41.409776, ymin = 51.344339), crs = st_crs(4326))

france <-st_crop(france, bbox_france)

ggplot() + 
  geom_sf(data=france, fill='white') +
  coord_sf()

proj_france <- 'PROJCS["ProjWiz_Custom_Lambert_Azimuthal",
 GEOGCS["GCS_WGS_1984",
  DATUM["D_WGS_1984",
   SPHEROID["WGS_1984",6378137.0,298.257223563]],
  PRIMEM["Greenwich",0.0],
  UNIT["Degree",0.0174532925199433]],
 PROJECTION["Lambert_Azimuthal_Equal_Area"],
 PARAMETER["False_Easting",0.0],
 PARAMETER["False_Northing",0.0],
 PARAMETER["Central_Meridian",1.8896484],
 PARAMETER["Latitude_Of_Origin",46.7708584],
 UNIT["Meter",1.0]]'
```

### 2.    Get 5 different drivers (`raster`) 

-   Temperature seasonality (**WorldClim BIO4**)
-   Precipitation seasonality (**WorldClim BIO15**)
-   Elevation (**SRTM**)
-   Water bodies (**ESA WorldCover**)
-   Human footprint index (**SEDAC**)

Extra:
-   Land-cover (**MODIS Terra MCD12Q1**) -> see demo online

```{r}
#| echo: true
#| eval: true

extent <- round(st_bbox(france),0) + c(-1,-1,1,1)

geodata_path("Practical_classes/Week3_predictors/data/")

bio <- worldclim_country(var = "bio", country = "FRA")

temp <- bio$wc2.1_30s_bio_4
temp <- crop(temp, extent)

prec <- bio$wc2.1_30s_bio_15
prec <- crop(prec, extent)

elev <- elevation_30s(country = "FRA")
elev <- crop(elev, extent)

land <- rast("Practical_classes/Week3_predictors/data/MCD12Q1.061_LC_Type1_doy2020001_aid0001_France.tif")
land <- crop(land , extent)

footp <- footprint(year = 2009)
footp <- crop(footp, extent)
```

### 3.    Process all the layers to the same resolution and the extent of France

```{r}
#| echo: true
#| eval: true

compareGeom(tavg, prec, elev, footp, land, stopOnError = F, messages = T, res = T)

r <- rast(res = 10000, # Target cell size
          ext(vect(st_transform(france, proj_france))), # Extent of the raster
          crs = proj_france) %>% # Useful planar projection
  project(., "epsg:4326") # Projection of our rasters

temp <- resample(temp, r, method = "bilinear") # for continuous data!
prec <- resample(prec, r, method = "bilinear")
elev <- resample(elev, r, method = "bilinear")
footp <- resample(footp, r, method = "near") # for categorical data!
land <- resample(land, r, method = "near") 

compareGeom(tavg, prec, elev, footp, land, stopOnError = F, messages = T, res = T)

temp <- mask(temp, vect(france))
prec <- mask(prec, vect(france))
elev <- mask(elev, vect(france))
footp <- mask(footp, vect(france))
land <- mask(land, vect(france))
```

### 4.    Visualise predictors 

```{r}
#| echo: true
#| eval: true

tmap_mode(mode = "view")

tm_shape(mean(temp)) +
  tm_raster(palette = "OrRd",
            midpoint = NA, 
            style = "cont", n = 10)

tm_shape(mean(prec)) +
  tm_raster(palette = "RdBu", 
            midpoint = NA, 
            style = "cont", n = 10) +
  tm_view(bbox = extent)

tm_shape(elev) +
  tm_raster(palette = "BrBG", 
            midpoint = NA, 
            style = "cont", n = 10) +
  tm_view(bbox = extent)

land <- as.factor(land$MCD12Q1.061_LC_Type1_doy2020001_aid0001_France)

levels(land[[1]]) <- data.frame(
  ID = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17),
  label = c(
    "Evergreen Needleleaf Forest",
    "Evergreen Broadleaf Forests",
    "Deciduous Needleleaf Forests",
    "Deciduous Broadleaf Forests",
    "Mixed Forests",
    "Closed Shrublands",
    "Open Shrublands",
    "Woody Savannas",
    "Savannas",
    "Grassslands",
    "Permanent Wetlands",
    "Croplands",
    "Urban and Built-up Lands",
    "Cropland/Natural Vegetation Mosaics",
    "Permanent Snow and Ice",
    "Barren",
    "Water Bodies"
  )
) 

tm_shape(land) +
  tm_raster(palette = "Set1", 
            style = "cat") +
  tm_view(bbox = extent) +
  tm_layout(legend.position = 'left')


tm_shape(footp) +
  tm_raster(palette = "viridis", style = "cont") +
  tm_view(bbox = extent)
```

### 5.    Calculate the correlation between the different predictors (and compare with biodiversity) 

Please download the following file and store it on you `data` folder:  

-   `trees_grid.rds` 

```{r}
#| echo: true
#| eval: true

predictors <- tibble(
  "tavg" = values(mean(temp)) %>% as.vector(),
  "prec" = values(mean(prec)) %>% as.vector(),
  "elev" = values(elev) %>% as.vector(),
  "land" = values(land) %>% as.vector(),
  "footp" = values(footp) %>% as.vector()
) %>% 
  mutate(
    across(everything(), ~replace_na(.x, NA))
  )


summary(predictors)

GGally::ggpairs(predictors, cex = 0.1)

trees_grid <- readRDS("Practical_classes/Week3_predictors/data/trees_grid.rds") %>% st_transform(., 4326) %>% st_intersection(france)

trees_grid <- trees_grid %>% st_intersection(france)

tm_shape(trees_grid) +
  tm_fill("SR", palette = "YlOrBr") +
  tm_view(bbox = extent)

```