---
title: "How to plot biodiversity data, explore patterns at different resolutions, and make pretty map"
authors: 
  -Florencia Grattarola
  -Friederike Woelke
  -Gabriel Ortega

format: docx
editor: visual
---

# Preparations before starting to code

## Spatial analyses

Install the package `sf`

```{r}
#| eval: false
install.packages('sf') # install
```

Load the library and check it's working.

```{r}
library(sf) # load
packageVersion('sf')
```

And switch spherical geometry off

```{r}
sf_use_s2(FALSE) # switch spherical geometry off
```

## Spatial download

For downloading country-specific and world polygons we will use the `rnaturalearth` package

```{r}
#| eval: false
install.packages(c('rnaturalearth', 'rnaturalearthdata')) # the second package is also needed
```

To use it, we load the library and check it's working.

```{r}
library(rnaturalearth) # load
packageVersion('rnaturalearth')
```

## Visualisation

For all the map plots today we will use the `ggplot` package. It's part of the `tidyverse`.

```{r}
library(tidyverse) # load
```

# Project variables

It is advisable to have a section of project/work variables in your scripts. That way it will be easy to reuse your scripts in other tasks.

## Create your project variables

Define here the things you already know you will use later on the script. For instance, we know that we will work with **mammals** from **Czechia** and the data we will get from GBIF are in **WGS84** latitude and longitude.

```{r}
taxa <- "Mammalia"
country_code <- "CZ" # Two letters ISO code for Czechia
proj_crs <- 4326 # EPSG code for WGS84
```

## Mapping occurrence records ðŸ¦Œ

1.  Get occurrence records of mammals in the Czech Republic (`POINTS`)

Get the data from last practical (downloaded from GBIF)

```{r}
mammalsCZ <- readRDS('data/mammalsCZ.rds')
mammalsCZ %>% head(n=10) %>% knitr::kable()
```

Let's keep only a few fields

```{r}
mammalsCZ <- mammalsCZ %>% select(species, order, eventDate, 
                                  decimalLongitude, decimalLatitude,
                                  stateProvince, countryCode, datasetName)
```

We will transform our data (table) into an sf object using `st_as_sf()`

```{r}
mammalsCZ %>% 
  filter(!is.na(decimalLongitude) & 
           !is.na(decimalLatitude)) %>% # filter records without coordinates
  st_as_sf(coords=c('decimalLongitude', 'decimalLatitude'),
           crs=4326) %>% head(n=10) %>% knitr::kable()
```

Note that `crs=4326` is related to the [EPSG:4326](https://epsg.io/4326) CRS: WGS 84.

2.  Create a grid layer of the Czech Republic (`POLYGONS`)

Load the administrative borders using `st_read`

```{r}
CZ_borders <- st_read('data/CZE_adm0.gpkg')
CZ_borders %>% head(n=10) %>% knitr::kable()
```

We will *actually* load a Czech standard grid layer of 100 km^2^ area

```{r}
CZ_grids <- st_read('data/KvadratyCR_JTSK.gpkg')
CZ_grids %>% head(n=10) %>% knitr::kable()
```

Check the layer's Coordinate Reference System (CRS)

```{r}
st_crs(CZ_borders) == st_crs(CZ_grids)

st_crs(CZ_borders)
st_crs(CZ_grids)
```

3.  Calculate sampling effort (N) and species richness per grid cell (SR)

Frist, we will transform the layer's CRS to [S-JTSK / Krovak East North](https://epsg.io/5514) using `st_transform()`.

```{r}
#| echo: true
#| eval: true
mammalsCZ_sf <- st_transform(mammalsCZ_sf,  
                             crs = st_crs(CZ_grids)) # the same CRS as the CZ_grids layer

mammalsCZ_sf %>% head(n=10) %>% knitr::kable()
```

Let's plot the sf objects

```{r}
#| echo: true
#| eval: true
ggplot() + 
  geom_sf(data=CZ_borders, fill='white') + # first layer is the borders with white fill
  geom_sf(data=CZ_grids, fill=NA) +  # first layer is the grid cells with no fill
  geom_sf(data=mammalsCZ_sf) # last layer is the points, that are black by default
```

To calculate number of records and number of species per grid cell, we will use `st_join()`

```{r}
#| echo: true
#| eval: true
st_join(CZ_grids,       # POLYGONS (grid cells)
        mammalsCZ_sf) %>%   # POINTS
   head(n=10) %>% knitr::kable()
```

Now, we need to summarise the data per grid-cell. Luckily, `tidyverse` methods also work for `sf` objects :) We will do it with `group_by()` and `summarise()`.\

First, we will group by grid-cells and then we count values per grid.
So, let's calculate the number of records (`N`) and the number of species per grid-cell (`SR`)

```{r}
#| echo: true
#| eval: true
st_join(CZ_grids, mammalsCZ_sf) %>% 
  group_by(OBJECTID) %>% # the name of the column that has the index
  summarise(N=sum(!is.na(species)), # calculates the number of points in the polygon
            SR=n_distinct(species, na.rm = TRUE)) %>%  # the number of different 'species' in the polygon
  head(n=10) %>% knitr::kable()
```

Let's store the output into a new object `CZ_mammals_grids`.

```{r}
#| echo: true
#| eval: true
mammalsCZ_grids <- st_join(CZ_grids, mammalsCZ_sf) %>% 
  group_by(OBJECTID) %>% 
  summarise(N=sum(!is.na(species)), 
            SR=n_distinct(species, na.rm = TRUE))
```

4.  Visualise **N** and **SR** patterns in Czech Republic

Finally, let's plot this. We will do it using `geom_sf()`, a `ggplot` function to visualise sf objects.  

```{r}
#| echo: true
#| eval: true
ggplot() + 
  geom_sf(data=mammalsCZ_grids) +
  coord_sf(crs=4326)
```

Where are the nice colours? We need to indicate which column from the object should the grids be filled with.

```{r}
#| echo: true
#| eval: true
ggplot() + 
  geom_sf(data=mammalsCZ_grids, aes(fill=SR)) +
  coord_sf(crs=4326)
```

Let's get a nicer color scale. Bear in mind that [we see colors differently](https://davidmathlogic.com/colorblind/). Thus, it's important to consider colorblind safe palettes.  
```{r}
#| echo: true
#| eval: true
ggplot() + 
  geom_sf(data=mammalsCZ_grids, aes(fill=SR)) +
  scale_fill_fermenter(palette ='YlOrBr', n.breaks=9, direction = 1) + # fill of the grids
  geom_sf(data=CZ_borders, fill=NA) +
  coord_sf(crs=4326) +
  theme_bw()
```

Now, here's a better plot :)

The hotspots of species richness are in cities. How can these be the richest areas for mammals?
Let's have a look at the sampling effort (`N`: number of records per grid cell) and compare both layers.  

```{r}
#| echo: true
#| eval: true
ggplot() + 
  geom_sf(data=mammalsCZ_grids, aes(fill=N)) +
  scale_fill_fermenter(palette ='YlGnBu', n.breaks=9, direction = 1) +
  geom_sf(data=CZ_borders, fill=NA) +
  coord_sf(crs=4326) +
  theme_bw()
```


What can you say about the hotspots of species richness we found?

```{r}
#| echo: false
#| eval: true
#| layout-ncol: 2
#| fig-cap: 
#|   - "Species richness"
#|   - "Number of records"

SR <- ggplot() + 
  geom_sf(data=mammalsCZ_grids, aes(fill=SR)) +
  scale_fill_fermenter(palette ='YlOrBr', n.breaks=9, direction = 1) +
  geom_sf(data=CZ_borders, fill=NA) +
  coord_sf(crs=4326) +
  theme_bw()

N <- ggplot() + 
  geom_sf(data=mammalsCZ_grids, aes(fill=N)) +
  scale_fill_fermenter(palette ='YlGnBu', n.breaks=9, direction = 1)+
  geom_sf(data=CZ_borders, fill=NA) +
  coord_sf(crs=4326) +
  theme_bw()

SR 
N
```

We are done! No it's **your turn** :)

## Mapping species ranges ðŸ¢

1.   Get testudines' IUCN expert range maps (`POLYGONS`)

Read the data (downloaded from IUCN [Spatial Data Download's](https://www.iucnredlist.org/resources/spatial-data-download) page). Take a look at the zip file for more info on metadata, and terms and conditions of use. 

```{r}
testudines <- st_read('data/testudines.gpkg')
```


What do the data look like?

```{r}
testudines %>% head(n=10) %>% knitr::kable()
```

2.   Get a world map

We will download a polygon of the world at medium scale and we will combine all countries into a unique polygon.


```{r}
world <- ne_countries(scale = 50, returnclass='sf')
world <- st_union(world) %>% st_make_valid() %>% st_cast() # fixes any problems
```

Let's see how it looks

```{r}
ggplot() + 
  geom_sf(data=world, fill='white')
```

Before working with this map, we need to **project** the layer: from lat/lon to equal area projection.
Earth is not flat :) **Projections** help us represent the two-dimensional curved surface of the globe into 2D space. There are many ways to do this. Find here a cool (Projection Wizard)[https://projectionwizard.org].

Equal-area maps preserve area measure, generally distorting shapes in order to do that

```{r}
#| echo: false
#| eval: true
#| layout-ncol: 3
#| fig-cap: 
#|   - "World Geodetic System (EPSG:4326)"
#|   - "Mollweide (ESRI:54009)"
#|   - "WGS 84 / Equal Earth Greenwich (EPSG:8857)"

ggplot() + 
    geom_sf(data=world, fill='white') + 
    coord_sf(crs = 4326)

ggplot() + 
    geom_sf(data=world, fill='white') + 
    coord_sf(crs = 'ESRI:54009')

ggplot() + 
    geom_sf(data=world, fill='white') + 
    coord_sf(crs = 'EPSG:8857')
```

We will choose **Equal Earth** (EPSG:8857).
Let's transform the data (both the world and the testudines' layers)

```{r}
world_ea <- st_transform(world, crs = 'EPSG:8857') %>% 
  st_make_valid() %>% st_cast()

testudines_ea <- st_transform(testudines, crs = 'EPSG:8857') %>% 
  st_make_valid() %>% st_cast()
```

Let's double check that everything is alright

```{r}
st_crs(world_ea, parameters = TRUE)$epsg
st_crs(world_ea, parameters = TRUE)$ud_unit
st_crs(testudines_ea, parameters = TRUE)$epsg
st_crs(testudines_ea, parameters = TRUE)$ud_unit
```

3.   Create a global 1-degree grid  (`POLYGONS`)

We will do this using the function `st_make_grid()`. For the `cellsize` argument we will chose 1 degree, which is \~100km (=100,000m)


```{r}
world_grid <- st_make_grid(world_ea, 
                           cellsize=100000,
                           square=FALSE) %>%  # this will make hexagons
  st_intersection(world_ea) %>% 
  st_sf(gridID=1:length(.))     
```

We have the grid, we are ready to calculate metrics per grid cell. Let's see how it looks.

```{r}
#| echo: false
#| eval: true
ggplot() + 
  geom_sf(data= world_grid, fill='white', size=0.1) 
```

4.   Calculate the species richness per grid cell (SR)

To do this, let's keep only the *testudines* that are extant (resident) species

```{r}
#| echo: true
#| eval: true
testudines_extant <- testudines_ea %>% 
  filter(LEGEND=='Extant (resident)') %>% 
  group_by(ID_NO) %>% 
  summarise(BINOMIAL=unique(BINOMIAL),
            PRESENCE=sum(PRESENCE)) %>% 
  ungroup() %>% st_cast()
```

This **can** take a while â˜•ï¸

Let's plot 10 species as an example

```{r}
ggplot() + 
  geom_sf(data= world_grid, fill='white', size=0.5)  +
  geom_sf(data= testudines_extant %>% head(n=10), aes(fill=BINOMIAL), alpha=0.5, lwd = 0)
```


Now let's use `st_join()` to calculate the number of species per grid-cell (SR).

```{r}
testudines_grid <- st_join(world_grid, testudines_extant) %>%
  group_by(gridID) %>%
  summarise(N=sum(!is.na(BINOMIAL)), 
            SR=n_distinct(BINOMIAL, na.rm = TRUE))
```

This **will** take a while â˜•ï¸ðŸª  

Here are the results :)

```{r}
#| echo: true
#| eval: true
testudines_grid %>% head(n=10) %>% knitr::kable()
```

5.   Visualise **SR** patterns in the world\

And now, we plot!

```{r}
ggplot() + 
  geom_sf(data=testudines_grid %>% mutate(SR=ifelse(SR==0, NA, SR)), aes(fill=SR), lwd = 0) +
  scale_fill_fermenter(palette ='YlOrBr', na.value ='grey90', n.breaks=6, direction = 1) +
  geom_sf(data=world_ea, fill=NA, col='grey70', size=0.2) +
  theme_bw()
```
