---
title: 'Downloading, cleaning and mapping data from GBIF'
date: 2025-09-29
author:
  - name: 'Florencia Grattarola'
    orcid: 0000-0001-8282-5732
format: 
  html:
    toc: true
    toc-location: right
    smooth-scroll: true
    html-math-method: katex
    code-fold: true
self-contained: true
editor: source
editor_options: 
  chunk_output_type: console
---

# Mammal's of the Czech Republic

We will use the mammals of the Czech Republic as an example dataset.
We will access the data through [**GBIF**](https://gbif.org) using R. 

## Some preparation before starting to code

  - Create a new project – you will use it for all your practical sessions (with `code` and `data` folders inside).  
  - Comment your code as much as possible, as if you were to explain it to others (that other could be you in 3 months!).  
  - Keep your code short and easily readable in plain English.

## 1 Install and load libraries

We will always load packages into R using the package **`pacman`**. It is a handy package for installing, loading, and unloading libraries. If you attempt to load a library that is not installed, `pacman` will try to install it automatically.

```{r}
#| eval: false
install.packages('pacman')
```

# 1 Install and/or load `tidyverse`

Tidyverse is a set of libraries for easier manipulation and transformation of data. We will be using many functions from this package, like `filter()`, `mutate()`, and later `read_csv()`.

```{r}
pacman::p_load(tidyverse) # Data wrangling
```

## 1.1 Install and/or load `rgbif`

We will use **`rgbif`** to download data from GBIF directly into our R session.

```{r}
#| echo: true
#| eval: true
 
pacman::p_load(rgbif) # the GBIF R package
```

## 1.2 Install and/or load `taxize`

We will need to get a taxon ID (`taxonKey`) for the *Mammalia* class from the GBIF backbone. For that, we will use another package called **`taxize`**.

```{r}
#| echo: true
#| eval: true
 
pacman::p_load(taxize)
```

## 1.3 Install and/or load `sf`

We will use **`sf`** to work with spatial data.

```{r}
#| echo: true
#| eval: true
 
pacman::p_load(sf)
```

## 1.4 Install and/or load `rnaturalearth`

We will use **`rnaturalearth`** to interact with [Natural Earth](https://www.naturalearthdata.com/) and get mapping data (e.g., countries' polygons) into R.

```{r}
#| echo: true
#| eval: true
 
pacman::p_load(rnaturalearth)
```

# 2 Project variables

It is advisable to have a section of project/work variables in your scripts. That way, it will be easy to reuse your scripts in other tasks.

## 2.1 Create a taxon ID for the *Mammalia* class

Define here the things you already know you will use later in the script. For instance, we know that we will work with **mammals** from the **Czech Republic** and that the data we will get from GBIF are in **WGS84** latitude and longitude.

```{r}
#| echo: true
#| eval: true
 
taxon <- 'Mammalia'
country_code <- 'CZ' # Two letters ISO code for the Czech Republic
proj_crs <- 4326 # EPSG code for WGS84
```

Let's get the taxon ID for the *Mammalia* class

```{r}
#| echo: true
#| eval: true
 
taxon_key <- taxize::get_gbifid_(taxon) %>%
  bind_rows() %>% # Transform the result of get_gbifid into a dataframe
  filter(matchtype == 'EXACT' & status == 'HIGHERRANK') %>% # Filter the dataframe by the columns 'matchtype' and 'status'
  pull(usagekey) # Pull the contents of the column 'usagekey'
```

## 2.2 Download a basemap of Czech Republic

It is often useful to have a base map of the study area (in this case the whole the Czech Republic).

```{r}
#| echo: true
#| eval: true
#| warning: false
#| message: false
 
base_map <- rnaturalearth::ne_countries(
  scale = 110,
  type = 'countries',
  country = 'Czechia',
  returnclass = 'sf'
)
```

# 3 GBIF data download

## 3.1 Check GBIF's data

Now, we can use the function `occ_count()` to find the **number of occurrence records** for the Czech Republic.

How many occurrence records are in GBIF for the entire **Czech Republic**?

```{r}
#| echo: true
#| eval: true

occ_count(country = country_code)
```

And how many of those records are **mammals**?

```{r}
#| echo: true
#| eval: true

occ_count(
  country = country_code,
  taxonKey = taxon_key
)
```

After this initial exploration, we are ready to download data. To do this, we will use `occ_search()`. This function has many options that correspond with fields in the GBIF database.

## 3.1 CZ mammals' GBIF data download

Get occurrences records of mammals from Czech Republic.

```{r}
#| echo: true
#| eval: true

occ_search(taxonKey = taxon_key,
           country = country_code) 
```

By default, it will only return the first 500 records. To get **all** the records, we need to specify a larger limit. Since we have over 15,000 records, we'll choose 16,000 as the limit. **However**, this will get super slow, so you could pick 5000.

```{r}
#| echo: true
#| eval: false
 
occ_search(taxonKey = taxon_key,
           country = country_code,
           limit = 5000) 
```

Finally, we store the result in the object `mammalsCZ`. We include the option to remove common geospatial issues (e.g., zero coordinates, country coordinate mismatch, invalid coordinate, etc.).  

```{r}
#| echo: true
#| eval: true
 
mammalsCZ <- occ_search(
  taxonKey = taxon_key, # Key 359 created previously
  country = country_code, # CZ, ISO code of the Czech Republic
  limit = 5000, # Max number of records to download
  hasGeospatialIssue = F # Only records without spatial issues
)

mammalsCZ <- mammalsCZ$data # The output of occ_search is a list with a data object inside. Here we pull the data out of the list.
```

# 4 Data exploration

Examine the dataset's variables and their respective data types: Are they `numeric`, `character`, or `boolean` in nature? 

```{r}
#| echo: true
#| eval: false
 
glimpse(mammalsCZ)
```

How many records do we have?

```{r}
#| echo: true
#| eval: true
 
nrow(mammalsCZ)
```

How many species do we have?

```{r}
#| echo: true
#| eval: true
 
mammalsCZ %>%
  filter(taxonRank == 'SPECIES') %>%
  distinct(scientificName) %>%
  nrow()
```

`distinct()` is used to see unique values

# 5 Data quality

Data are not 'good' or 'bad', the quality will depend on our goal. Some things we can check:

-   Basis of the record - the specific nature of the record (type of occurrence)  
-   Species names (taxonomic harmonisation)  
-   Spatial and temporal (accuracy / precision)  

## 5.1 Basic data filtering

As an example of data cleaning procedures, we will check the following fields in our dataset:

-   `basisOfRecord`: we want preserved specimens or observations
-   `taxonRank`: we want records at species level.
-   `coordinateUncertaintyInMeters`: we want them to be smaller than 10km.

### `basisOfRecord`:

We want preserved specimens and observations

```{r}
#| echo: true
#| eval: true
 
mammalsCZ %>% distinct(basisOfRecord)
```

Count how many records are in each type of *basisOfRecord*

```{r}
#| echo: true
#| eval: true
 
mammalsCZ %>%
  group_by(basisOfRecord) %>%
  count()
```

`group_by()` is used to group values within a variable

Update the object by filtering over the *basisOfRecord* to keep only records that correspond to 'preserved specimens' or 'human observations'.

```{r}
#| echo: true
#| eval: true
 
mammalsCZ <- mammalsCZ %>%
  filter(basisOfRecord == 'PRESERVED_SPECIMEN' |
    basisOfRecord == 'HUMAN_OBSERVATION')
```

Note the use of `|` (OR) to filter the data. Another alternative is `filter(basisOfRecord %in% c('PRESERVED_SPECIMEN','HUMAN_OBSERVATION'))`

How many records do we have now?

```{r}
#| echo: true
#| eval: true
 
nrow(mammalsCZ)
```

### `taxonRank`:

We want records at species level

```{r}
#| echo: true
#| eval: true
 
mammalsCZ %>% distinct(taxonRank)
```

Update the object by filtering over **taxonRank** to keep only records that correspond to the 'species' level. 

```{r}
#| echo: true
#| eval: true
 
mammalsCZ <- mammalsCZ %>%
  filter(taxonRank == 'SPECIES')
```

How many records do we have now?

```{r}
#| echo: true
#| eval: true
 
nrow(mammalsCZ)
```

### `coordinateUncertaintyInMeters`:

We want uncertainty to be smaller than 10km.

Are there records with precision larger than 10km?

```{r}
#| echo: true
#| eval: true
 
mammalsCZ %>%
  filter(coordinateUncertaintyInMeters >= 10000) %>%
  select(scientificName, 
         coordinateUncertaintyInMeters, 
         stateProvince) 
```

Update the object by filtering over **coordinateUncertaintyInMeters** to keep only records that have less than 10000 meters of uncertainty. 

```{r}
#| echo: true
#| eval: true
 
mammalsCZ <- mammalsCZ %>%
  filter(coordinateUncertaintyInMeters < 10000) # keeping this
```

How many records do we have now?

```{r}
#| echo: true
#| eval: true
 
nrow(mammalsCZ)
```

# 6 Basic maps

How are the records distributed?

```{r}
#| echo: true
#| eval: true
 
library(sf)

mammalsCZ_sf <- mammalsCZ %>%
  filter(!is.na(decimalLongitude) &
    !is.na(decimalLatitude)) %>%
  st_as_sf(coords = c(
    'decimalLongitude',
    'decimalLatitude'
  )) %>%
  st_set_crs(proj_crs)

ggplot() +
  geom_sf(data = base_map, fill = 'white') +
  geom_sf(
    data = sf::st_intersection(mammalsCZ_sf, base_map),
    aes(col = order)
  ) +
  theme_bw()
```

And finally, a simple trick to produce separate maps per order.

```{r}
#| echo: true
#| eval: true
 
ggplot() +
  geom_sf(data = base_map, fill = 'white') +
  geom_sf(
    data = sf::st_intersection(mammalsCZ_sf, base_map),
    aes(col = order)
  ) +
  theme_bw() +
  # Added to the previous ggplot
  facet_wrap(~order) +
  theme(
    legend.position = 'none',
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
  )
```

# EXTRA

Now, please create an account in GBIF for our next practical and follow the script [download_mammalsCZ_data_from_GBIF.R](https://github.com/petrkeil/Spatial-Ecology-and-Macroecology/blob/main/Practical_classes/Week2_gridding_and%20plotting/download_mammalsCZ_data_from_GBIF.R) in the **Week2_gridding_and plotting** of the **Practical_classes** folder to get these data. 


