---
title: "Practical 8: Evolution of Biodiversity & Diversification"
author: "Friederike Wölke, wolke@fzp.czu.cz, Room D330"
format:
  html:
    theme: united
    toc: true
    code-overflow: wrap
    embed-resources: true
---

This document outlines steps for **analyzing mammalian biodiversity using phylogenetic trees and diversification** **metrics**. It's part 1 of [todays' practical on evolution of biodiversity](https://github.com/petrkeil/Spatial-Ecology-and-Macroecology/tree/main/Practical_classes/Week8_evolution_of_biodiversity) in [**spatial ecology and macroecology 2024.**](https://github.com/petrkeil/Spatial-Ecology-and-Macroecology)

I have prepared a config.R file that has all the packages that we are going to need for today. Since it's a lot (on my machine at home it took less than 5 minutes, but you never know with the computers at Uni) we will run it in parallel while you already have a look at the prepared code/outputs in the .html file.

**At the bottom of this file you will find an activity where you chose your favorite mammal clade, run a diversification analysis, and plot the results on the world map.**

Please ensure that all packages are installed with the following code chunk to being able to do this activity.

::: {.callout-tip icon="false"}
## ⭐ A note on frustration:

If it doesn't work and you don't want to make it work: it's fine. The topic of evolution is a bit out of scope for the course anyway. If you can't make it work but want to make it work, please let me know and we will find a solution.
:::

## Prepare R

```{r}
#| label: configurations

source("config.R")

```

# Case study: Mammals in Botanical Countries

## Setting up path variables

Here we are saving the paths to our files as character strings to have a cleaner (and more adaptable) script downstream. It's recommended to set up variables in the beginning of a script to keep things clean and easy to change.

**What we need:**

1.  The phylogenetic tree (in nexus format)
2.  Mammal occurrence data (prepared using the script 'preprocess_mammals_ranges.R'
3.  Taxonomic information for subsetting and annotation
4.  The shapefile in botanical countries-scale (tdwg3).

**Thanks to the [PHYLACINE](https://github.com/MegaPast2Future/PHYLACINE_1.2/) database, we have it all together already :-)**

Familiarize yourself with the `here-package`. It's the new reproducible and collaborative way to do R stuff on different machines. It always sets the working directory to be where the script is/or the Rproj. if you have one.

```{r, libraries}
#| label: set-path-vars

# Clean working space
rm(list = ls())

# Paths to our data:

out_path <- here("data") # where we will save stuff

tree_path <- here("data_raw", "Phylogenies", "Complete_phylogeny.nex") # file path to the phylogeny (in nexus format)

occurrence_path <- here("data", "mammals_occu.rds") # file path to the occurrence data that I prepared with the 'preprocess_mammals_ranges.R' script.

tax_path <- here("data_raw", "Traits", "Trait_data.csv") # Taxonomic information from the trait database

shp_path <- here("data_raw","shp") # shapefile in botanical countries-scale


# Let's set the theme for ggplots
theme_set(theme_bw())

```

## The Data

[Source:]{.underline} <https://github.com/MegaPast2Future/PHYLACINE_1.2/>

### 1) Phylogenetic tree for mammals

::: {.callout-tip icon="false"}
## ⭐ A note on Phylogenetic Trees

Phylogenetic trees are very uncertain and should be regarded as *'hypothesis of relationship'* rather than facts. They are created using bayesian methods and the output of such analyses is a posterior distribution of the result. Therefore, they are very rarely provided as a single tree and rather as a set of 1000 phylogenetic trees that were sampled randomly from said posterior distribution. These trees vary in placement of clades within the tree and branching times. This makes phylogenetic analyses require a lot of computational power, since one line of code can easily take a week to run for a set of 1000 phylogenetic trees with sufficient size.

We will therefore find the *maximum clade credibility* (MCC) tree, which tries to find the tree with the highest confidence in the placement of clades from the set of 1000 trees.

[***Please skip this code chunk right below and read in the file for the MCC tree instead (next chunk).***]{.underline}
:::

```{r, phylogeny, eval = FALSE}
#| eval: false
#| label: create-mcc-phylogeny

### PLEASE DO NOT RUN ###

tree <- read.nexus(tree_path)

# The Maximum Clade Credibility tree: 
# (this takes about 1h to calculate. I did it for you! Skip to the next code block)
mcc_tree <- mcc(tree, tree = TRUE, rooted = TRUE)

# Inspect the object:
mcc_tree
saveRDS(mcc_tree, here(out_path, "MCCtree.rds"))

#### END #####

rm(mcc_tree, tree, tree_path)

```

```{r}
#| label: read-phylogeny

### RUN THIS INSTEAD ###

tree <- readRDS(file.path(out_path, "MCCtree.rds")) %>% 
  # this will make the tree look more organized but doesn't change its structure:
  ladderize() 

```

#### 🖼️Phylogeny plot

::: callout-note
## 🗨️Discussion 1

What can we say about diversification in Mammals?

-   when? (in which period?)

-   how strong?

-   any trends?
:::

```{r}
#| label: plot-phylogeny
par(mfrow = c(1,1))
plotTree(
  tree,   
  ftype = "off", 
  ylim=c(-0.2,1)*Ntip(tree)) # increase the plot window to the bottom to plot the geo axis
# add the axis
phytools::geo.legend()

```

#### 🖼️Lineages through time plot

Let's visualize the diversification of lineages (i.e, species) through time, and indicate when the *K-Pg Mass Extinction* happened 66 million years ago. Let's have a look at the temporal dynamics of diversification.

::: callout-note
## 🗨️Discussion 2

-   Identify the direction of the x-axis: Is it from past to present or from present to past?

-   How old is the mammal clade? When did it evolve?

-   When did mammals diversification get started?

-   Compare the two visualizations: why does it make sense to use the log-transformation?
:::

```{r}
#| label: lineages_through_time_plot

par(mar = c(5, 4, 4, 2) + 0.1) # set up the plot window


# Lineages-through-time plot

K_P_massextinction <- 217-66

ltt1 <- ltt(tree, log.lineages = FALSE, plot = FALSE)

par(mfrow=c(1,2))
plot(ltt1, log.lineages = FALSE, main = "# Lineages");abline(v = K_P_massextinction, col = "red")
plot(ltt1, log.lineages = TRUE, main = "log # Lineages");abline(v = K_P_massextinction, col = "red")
par(mfrow=c(1,1))

rm(ltt1) # let's remove it because it's bigg and we don't need it anymore
```

### 2) Occurrence data for mammals (country-scale)

I prepared mammal occurrences for you beforehand, because it takes some time for almost 6000 species. I have provided the code how I prepared this data in the project folder. The range maps were taken from the PHYLACINE database and can be freely downloaded from their repository to reproduce the occurrence data I provided here.

For the sake of time, we will be using the data that I have prepared instead of re-doing the preprocessing.

```{r, occurrence-data}
#| label: read-occurrence-data

occur <- read.csv(here("data", "occ_data.csv")) 

# Inspect the data
class(occur) # sf object
head(occur) # presence of species in polygons (botanical countries/LEVEL_3_CO)
```

#### 🖼️TDWG3 world plot

::: callout-note
## 🗨️ Discussion 3

-   What can you see in the map? Describe the scale at which we will assess diversification.
:::

```{r}
#| label: viz-occu-data

# Fix some geometry issues (:
occur <- st_read(shp_path) %>%
  left_join(occur) 

# Visualize the map and its resolution
occur %>% 
  select(LEVEL_3_CO) %>% 
  unique() %>%
  ggplot()+
  geom_sf(aes(fill = LEVEL_3_CO), show.legend = FALSE)
```

Cool! Now we are prepped with our data and can calculate some **phylogenetic indices** to investigate global patterns for.

## Phylogenetic calculations

We start by setting up some vectors again.

::: {.callout-tip icon="false"}
## ⭐ A note on phylogenetic comparative methods

Most importantly, to run any phylogenetic analysis, the **species in the tree have to match the species in the data**. Not only that it has to be exactly the same species list, it's also needed for many analyses that the species lists are in the same order. Since the tree always has an internal structure that dictates the order of tip labels, it is advised to **reorder the data by tips in the tree.**
:::

```{r}
#| label: species-names-vectors

## Species vectors for matching 

sp_in_tree <- tree$tip.label
sp_in_data <- unique(occur$species)

```

### Calculate phylogenetic metrics (on the whole tree)

Once we have reconstructed the geographic pattern of species diversity, we can reconstruct the pattern of diversification. We will do it twice, once for present-day diversification rates (DR) and for whole-clade or deep-time diversification rates (RD).

**Present-day diversification (DR)** captures the diversification rates near the tips of the phylogeny (\~ last 2 myrs). DR can be be calculated as the inverse of **species distinctiveness (ED)**, which is a measure of species isolation on the phylogeny. Species positioned at the end of long branches are viewed as distinctive (high values of ED). It can be shown mathematically that these species are also associated with slow diversification (low values of DR). Fast diversification, conversely, typifies species with low ED. The inverse relationships between ED and DR (**DR = 1 / ED**) is rather intuitive, when you consider that rapid radiations produce many species on the phylogeny separated by short branches.

Here we compare these patterns to **deep-time diversification rates (root distance, RD)**: This one is a bit more inclusive towards diversification phases over the whole evolutionary time of the phylogeny. It counts the number of nodes (ancestors) between the species at the tip and the root. Higher numbers indicate higher diversification rates (over the whole clade, since its origin at the root).

```{r}
#| label: do-phylo-calculations


# Species evolutionary distinctiveness (ED) ## took 13 seconds
## there is a function for it from the picante-package.
ed <- evol.distinct(tree) %>%
  filter(Species %in% sp_in_data)

# Present-day diversification rates (DR)
dr <- 1 / ed$w
names(dr) <- ed$Species


# Diversification rates of whole clades (RD) ## took 57 seconds
## there is a function for it from the adephylo-package.
rd0 <- distRoot(tree, 
               tips = "all",
               method = c("nNodes")) 


# Now we filter for species for which we actually have data. 
rd <- rd0[names(rd0) %in% sp_in_data]

```

### Match tree and data

Now let's do the filtering and only keep tips that are found in both (tree and data).

```{r}
#| label: drop-tips-from-tree

# Drop species from the tree for which we do not have data
tree <- keep.tip(tree, 
                 intersect(sp_in_tree, sp_in_data)
                 ) %>% 
  # Ladderize the tree again
  ladderize()

```

#### 🖼️ Plot Diversification rates on the tree

The phylogeny can help to demonstrate what exactly DR measures, given that short branches are generally associated with high DR, while the opposite holds for long branches. Let's visualize it.

```{r}
#| label: plot-rates-on-tree
par(mfrow=c(1,2))

plotTree.wBars(tree, 
               dr, 
               type="fan",
               lwd = 0.2)


plotTree.wBars(tree, 
               rd, 
               type="fan",
               method = "plotTree",
               lwd = 0.2)

```

### Species richness

We will sum up all species in each botanical country to get the species richness.

```{r}
#| label: create-plotting-data-SR
# NOTE: if some countries are missing in the map, try uncommenting 'st_make_valid()'

richness <- occur %>% 
  group_by(LEVEL_3_CO) %>% 
  mutate(
    SR = sum(presence) # Calculate species richness per country:
    ) %>% 
  dplyr::select(LEVEL_3_CO, SR) %>%
  unique()

```

### Diversification Metrics

We will calculate mean diversification rates for each botanical country. Since the differences between different lineages can be large, we will scale the rates to be between 0 and 1.

```{r}
#| label: create-plotting-data-DR-RD

  
# Present-day diversification rates (DR)
dr_df <- as.data.frame(dr)
dr_df$species <- row.names(dr_df)
dr_df2 <- left_join(occur, dr_df)  # merge with occurrence data

diversification <- dr_df2 %>%
  filter(!is.na(dr)) %>%
  dplyr::select(LEVEL_3_CO, dr) %>%
  group_by(LEVEL_3_CO) %>%
  dplyr::mutate(
    mean_DR = mean(dr, na.rm = T)
    ) %>%
  dplyr::select(-dr) %>%
  unique() %>%
  ungroup() %>%
  mutate(
    scaled_DR = ((mean_DR - min(mean_DR)) / (max(mean_DR) - min(mean_DR)))
    )



# Clade diversification (Root distance, RD)
rd_df <- as.data.frame(rd)
rd_df$species <- row.names(rd_df)
rd_df2 <- left_join(occur, rd_df) # merge with occurrence data

root_dist <- rd_df2 %>%
  filter(!is.na(rd)) %>%
  dplyr::select(LEVEL_3_CO, rd) %>%
  group_by(LEVEL_3_CO) %>%
  dplyr::mutate(
    mean_RD = mean(rd, na.rm = T)
    ) %>%
  dplyr::select(-rd) %>%
  unique() %>%
  ungroup() %>%
  mutate(
    scaled_RD = ((mean_RD - min(mean_RD)) / (max(mean_RD) - min(mean_RD)))
    )

```

::: callout-note
## 🗨️ Discussion 4

-   Compare the maps: are there geographical trends/patterns?

-   Think of extinctions (in space): what do you think how did they influence this result?

-   Think of plate tectonics: what do you think is the reason for this pattern?
:::

#### 🖼️ Species richness *versus* Diversification

```{r}
#| label: plot-maps
tmap_options(check.and.fix = TRUE) # add this and ignore the warnings :-)

my_palette <- viridis(12, option = "D")  

# Species Richness today:
map1 <- tm_shape(richness) +
  tm_polygons("SR", palette = my_palette) + 
  tm_layout(main.title = "Mammals - Species Richness",
            legend.position = c("right", "bottom"))

# Recent diversifications:
map2 <- tm_shape(diversification) +
  tm_polygons("scaled_DR", palette = my_palette) + 
  tm_layout(main.title = "Mammals - Present-day Diversification",
            legend.position = c("right", "bottom"))

# Overall clade diversifications:
map3 <- tm_shape(root_dist) +
  tm_polygons("scaled_RD", palette = my_palette) + 
  tm_layout(main.title = "Mammals - Clade Diversification",
            legend.position = c("right", "bottom"))

tmap_arrange(map1, map2, map3, nrow = 3)
```

# ❇️Activity: Subclades - Your turn!

Let's load the taxonomic information from PHYLACINE and subset the data based on our interest.

This is just something I do to prepare the data for your task :)

```{r}
#| label: create-data-subsets

## Here I split the phylogeny into smaller clades (orders).
## If you want to investigate a  family or even the species within a genus, you can use this code template to filter your tree based on your interest :)


# Read the taxonomic information from Phylacine:
tax <- read.csv(tax_path)[1:3]  %>% 
  filter(Binomial.1.2 %in% sp_in_tree)


# Which one do you want to investigate? 
orders <- unique(tax$Order.1.2)
family <- unique(tax$Family.1.2)
genus <- unique(tax$Genus.1.2)

# Define the Most Recent Common Ancestor (MRCA) function
fmrca = geiger:::.mrca  

# make clade lists
clades <- list()

## Orders: 
## OR Change the loop-vector to 'family' or 'genus' (whatever you prefer)

for (i in orders) { 
  print(i)
  
  # Subset tax based on the current order
  x <- subset(tax, Order.1.2 == i)
  
  # Find common labels between x$Binomial.1.2 and tree$tip.label
  valid_labels <- intersect(x$Binomial.1.2, tree$tip.label)
  
  # Proceed only if there are valid labels
  if (length(valid_labels) > 1) {  # MRCA requires at least 2 labels
    y <- extract.clade(tree, getMRCA(tree, valid_labels))
    clades[[i]] <- y
  } else {
    next  # Skip if no valid labels or only one label
  }
}


class(clades) <- "multiPhylo"

```

#### 🖼️ Orders Phylogeny plots

```{r}

# Set up a 5x5 plotting layout
par(mfrow = c(5, 5))  # Adjust margins as needed

# Loop over each tree in the 'clades' list and plot it
for (i in 1:length(clades)) {
  plotTree(clades[[i]], ftype="off", mar = c(2, 2, 2, 2))  
  title(main = paste(names(clades)[i]))
}
```

In the following code chunk, I have split the phylogeny by taxonomic orders. We have 25 different orders of mammals in this tree. **The goal is that you chose any one of these clades and investigate the geographic patterns of species richness & diversification rates.**

I'll run one example for Carnivores (because the clade is rather small but big enough and can be nicely visualized)

## Step 1: Chose your clade:

```{r}
#| label: Task-Subclades


## Adjust your clade of choice here!!
my_clade <- clades$Carnivora

par(mfrow = c(1,1))
# Plot your clade
plotTree(ladderize(my_clade), 
         fsize = 0.4, 
         ftype = "i", 
         lwd = 0.2, 
         type = "fan")

# extract species names
carni_sp <- my_clade$tip.label

# extract occurrence data & calculate SR:
carni_occ <- occur %>% 
  st_drop_geometry() %>% 
  filter(species %in% carni_sp) %>%
  group_by(LEVEL_3_CO) %>%
  mutate(
    SR_carni = n_distinct(species)
    ) %>% 
  distinct(SR_carni, LEVEL_3_CO)

# Spatial object:
carni_occ2 <- occur %>% 
  dplyr::select(LEVEL_3_CO) %>% 
  # Merge with Species Richness
  left_join(carni_occ) %>% 
  dplyr::select(LEVEL_3_CO, SR_carni) %>% 
  unique()

# Carnivores richness map =============
par(mar = c(5, 4, 4, 2) + 0.1) # set up the plot window
plot(carni_occ2["SR_carni"], main = "Species Richness - Carnivores")
ltt(my_clade, log.lineages = TRUE)

```

## Step 2: Calculate Diversification rates

```{r}
# From above: diversification rates =====
dr_df <- as.data.frame(dr)
dr_df$species <- row.names(dr_df)
dr_df2 <- left_join(occur, dr_df) 
##

div_carni <- dr_df2 %>% 
  # Filter here:
  filter(!is.na(dr) & species %in% carni_sp) %>%
  
  dplyr::select(LEVEL_3_CO, dr) %>%
  group_by(LEVEL_3_CO) %>% 
  dplyr::mutate(mean_DR = mean(dr, na.rm=T)) %>%
  dplyr::select(-dr) %>%
  unique() %>%
  ungroup() %>%
  mutate(scaled_DR =((mean_DR - min(mean_DR)) / (max(mean_DR) - min(mean_DR)))) %>%
  unique()

# We removed some countries by filtering before (those where SR mammals = 0 in this dataset)
div_carni2 <- occur %>% 
  dplyr::select(LEVEL_3_CO) %>% 
  left_join(div_carni %>% st_drop_geometry()) %>% 
  dplyr::select(LEVEL_3_CO, mean_DR, scaled_DR) %>% 
  unique()



### Root distance ======

rd_df <- as.data.frame(rd) 
rd_df$species <- row.names(rd_df)
rd_df2 <- left_join(occur, rd_df)

root_dist <- rd_df2 %>% 
  # Filter here:
  filter(!is.na(rd) & species %in% carni_sp) %>%
  
  dplyr::select(LEVEL_3_CO, rd) %>%
  group_by(LEVEL_3_CO) %>% 
  dplyr::mutate(mean_RD = mean(rd, na.rm=T)) %>%
  dplyr::select(-rd) %>%
  unique() %>%
  ungroup() %>%
  mutate(scaled_RD =((mean_RD - min(mean_RD)) / (max(mean_RD) - min(mean_RD))))

# We removed some countries by filtering before (those where SR mammals = 0 in this dataset)
root_dist_carni2 <- occur %>% 
  dplyr::select(LEVEL_3_CO) %>% 
  left_join(root_dist %>% st_drop_geometry()) %>% 
  dplyr::select(LEVEL_3_CO, mean_RD, scaled_RD) %>% 
  unique()


# Carnivores diversification map1 =============
plot(div_carni2["scaled_DR"], main = "Present-day diversification - Carnivores")

# Carnivores diversification map2 =============
plot(root_dist_carni2["scaled_RD"], main = "Clade diversification - Carnivores")

```

## Step3: Plot Richness vs. Diversification

```{r}
#Plot:
tm_shape(div_carni2) +
  tm_polygons("mean_DR", palette = my_palette, colorNA = "grey") + 
  tm_layout(main.title = "Carnivores - Present Day Diversification") +
  tm_layout(legend.outside = TRUE)

tm_shape(root_dist_carni2) +
  tm_polygons("mean_RD", palette = my_palette, colorNA = "grey") + 
  tm_layout(main.title = "Carnivores - Clade Diversification") +
  tm_layout(legend.outside = TRUE)

tm_shape(carni_occ2) +
  tm_polygons("SR_carni", palette = my_palette, colorNA = "grey") + 
  tm_layout(main.title = "Carnivores - Species Richness") +
  tm_layout(legend.outside = TRUE)

# Relationship between richness and diversification
plot(div_carni2$mean_DR, carni_occ2$SR_carni, "p")
```

# 🧚‍♂️🪄EXTRA

## Plotting tools: annotation

```{r}
#| label: tools-annotation

# create a color vector for plottin the labels on the tree:
# Orders: (for visualization on the full tree)
palette_28 <- brewer.pal(12, "Set3")  # Set3 has 12 colors by default
colors28 <- colorRampPalette(palette_28)(28)



# First, plot the tree without tip labels
plotTree(tree, type = "fan", ftype = "off")

# Initialize an empty list to store clades
clades <- list()

# Iterate over all orders
for (i in seq_along(orders)) {  # Correct the loop to iterate over orders properly
  
  # Get the current order label
  label <- orders[i]
  
  # Subset tax based on the current order
  x <- subset(tax, Order.1.2 == label)
  
  # Find common labels between x$Binomial.1.2 and tree$tip.label
  valid_labels <- intersect(x$Binomial.1.2, tree$tip.label)
  
  # Check if there are valid labels, and based on their number, adjust label properties
  
  if (length(valid_labels) > 20) {  
    
    # For clades with more than 15 valid labels
    # Extract the clade and save it to the list
    
    y <- extract.clade(tree, getMRCA(tree, valid_labels))
    clades[[label]] <- y  # Use label as the key for storing clades
    
    # Annotate the clade in the plot with curved labels
    arc.cladelabels(
      tree = tree, 
      text = label,  # Text to display
      node = findMRCA(tree, valid_labels),  # Node number to label
      orientation = "curved",  # Curve the label for fan plot
      col = colors28[i],  # Color of the label
      ln.offset = 1.05,  # Adjust label position to avoid overlap with the tree
      lab.offset = 1.10,
      cex = 0.5,  # Font size
      lwd = 3    # Line width for the label arc
    )
    
  } else if (length(valid_labels) <= 20 & length(valid_labels) > 1) { 
    
    # For clades with 2-15 valid labels
    # Annotate the clade in the plot with horizontal labels
    
    arc.cladelabels(
      tree = tree, 
      text = label,  # Text to display
      node = findMRCA(tree, valid_labels),  # Node number to label
      orientation = "horizontal",  # Horizontal for smaller clades
      col = colors28[i],  # Color of the label
      ln.offset = 1.05,  # Adjust line position to avoid overlap with the tree
      lab.offset = 1.15, # Adjust label position to avoid overlap with others 
      cex = 0.5,  # Font size
      lwd = 3    # Line width for the label arc
    ) 
  }
}

# Convert the clades list to a "multiPhylo" object
class(clades) <- "multiPhylo"


```

```{r}
#| eval: false
save.image(here(".RData"))
```
